Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2020-07-22 manuscript/v2_medrxiv/manuscript.tex
fix bibfile, compile tex

\section{Introduction}\label{introduction}

@@

\section{Causal Model}\label{causal-model}

@@

A number of studies report a strong association of \textbf{mobility}
restrictions on the number of new COVID-19 cases: Restrictive measures
(e.g. ``stay-at-home'' orders, travel bans, or school closures) are
shown to possibly reduce the COVID-19 incidence
\citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}.
However, some studies point out the combination of various
non-pharmaceutical interventions (NPIs) is decisive to prevent new
infections \citep{juni_impact_2020, lai_effect_2020}.

Google Trends \citep{google_trends} data can be used as a tool to get
insights into public interest (\textbf{awareness}) in the coronavirus
disease. Several recent studies imply a connection of relative search
volumes (RSV) indices and reported new COVID-19 cases
\citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}.
Some search terms e.g. ``COVID-19'' or ``coronavirus'' predated newly
infected cases/total number of cases by roughly 7 to 14 days for
different countries
\citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}.
Additionally, we acknowledged that individual risk-aware behavior might
be a reaction to the current COVID-19 burden (measured as reported cases
at the day of exposure).

@@

When investigating causal determinants of SARS-CoV-2 infections, a
number of confounders have to be considered. Well-known risk factors for
SARS-CoV-2 as well as for other infections are demographic factors such
as age, gender, socio-economic status (SES), population density, and
foreign citizenship/ethnicity
\citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In
Germany along with other countries (i.e.~Brazil, USA, or the UK),
populist parties or politicians and their electorate tend to be more
sceptical about effects of containment measures than the other part of
the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}.
Therefore we considered both ``right-wing populist party votes'' and
``voter turnout'' as possible confounders. Public health interventions
were also taken into account (contact restrictions, school closures
etc.), as their implementation showed strong correlations with
controlling the spread of SARS-CoV-2
\citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid
bias due to reporting delay of case numbers we had to include weekday
and German holidays. We include some unobserved variables in our DAG
(e.g. ``Herd immunity''), too. Please note that ``Exposure to
SARS-CoV-2'' is itself an unobserved variable: German case numbers are
reported with delay after date of exposure and symptom onset.
\emph{Exposure to the virus} should not be confused with the formal
\emph{exposure variables} of the DAG (mobility, awareness, temperature).

\section{Data}\label{data}

@@

\subsection{Variables}\label{variables}

@@

\section{Methods}\label{methods}

\subsection{Causal analysis with DAG and adjustment
sets}\label{causal-analysis-with-dag-and-adjustment-sets}

@@

\subsection{Regression with negative binomial
model}\label{regression-with-negative-binomial-model}

We can estimate the causal effect from exposure to outcome by regression
\citep{pearl2009causality}. Since the outcome ``Reported new cases of
COVID-19'' is a count variable, one should not employ a linear
regression model with Gaussian errors, but instead we assumed a
log-linear relationship between the expected value of the outcome \(Y\)
(new cases) and regressors \(x\), as well as a Poisson or negative
binomial distribution for \(Y\):

\begin{equation}
\log(\mathbb{E}[Y|x]) = \alpha + \sum_{i\in S}\beta_i\cdot x_{i}, \label{eq:regmodel}
\end{equation}

where \(\alpha\) is the regression intercept, \(S\) is the set of
adjustment variables for the exposure \(i^{\ast}\) including the
exposure variable itself, \(\beta_i\) are the regression coefficients
corresponding to the variables \(x_i\). As such \(\beta_{i^{\ast}}\) is
the total causal effect from exposure variable \(x_{i^{\ast}}\) on the
outcome Y.

@@

We needed to account for the fact that our outcome is not counted per
time unit (one day) only, but depends on the number of active COVID-19
cases: Holding all other variables fixed, the number of new cases \(Y\)
is a constant proportion of the number of active cases \(A\). This was
modeled by including an offset \(\log(A+1)\) in the regression model
\eqref{eq:regmodel}:

\begin{align}
\log(\mathbb{E}[Y|x]) &= \alpha + \log(A+1) + \sum_{i\in S}\beta_i\cdot x_{i} \nonumber \\
\Leftrightarrow \log(\frac{\mathbb{E}[Y|x]}{A+1}) &= \alpha + \sum\beta_i\cdot x_{i} \label{eq:logratemodel} \\
\Leftrightarrow \frac{\mathbb{E}[Y|x]}{A+1} &= \exp(\alpha)\cdot \prod \exp(\beta_i)^{x_i}. \label{eq:ratechange}
\end{align}

Here we added a pseudocount ``\(+1\)'' to ensure a finite logarithm and
avoid division by \(0\).

@@

\section{Results}\label{results}

@@

\subsection{Main results}\label{main-results}

@@

Within our framework, we saw strong effects for mobility in
retail/recreational areas and essential shopping (grocery and pharmacy).
In the former, an increase of 1 percent point mobility compared to the
reference period (03 January to 06 February 2020) leads to an increase
of the daily reported case number by about 0.8 percent. Contrarily, a
corresponding increase of 1 percent point for the areas of
grocery/pharmacy leads to a decrease in the reported case number by
approximately 0.5 percent. Mobility on workplaces showed a small effect
of 0.3 increase in case numbers for every 1 percent point increase in
mobility. Other causal effects of mobility were insubstantial and not
consistent in their direction (99 percent confidence intervals of
estimates include 1). Figure \ref{fig:effectsrange} shows the effects of
mobility on a range of possible values. Thus, we expect an increase of
daily cases by approximately 23 percent if mobility in retail/recreation
reaches baseline levels of 0 percent difference to the reference period.
On the other hand, an increase of mobility for grocery/pharmacy by 10
percent points compared to the reference period leads to a reduction of
the infection rate by approximately 7 percent.

@@

Within our model, we observed a causal effect of temperature and all
other weather variables. Every increase of 1 degree Celsius in
temperature leads to a reduction of the daily reported case numbers by
approximately 0.8 percent. Similarly strong was the effect of rainfall:
One millimeter (=1 liter per square meter) more rainfall leads to an
increase of reported case numbers by approximately 1.1 percent. We
observe small effects for humidity and wind as well (higher humidity
leading to fewer cases, stronger wind leading to less cases). In
perspective (Figure \ref{fig:effectsrange}), with temperature we expect
an increase by approximately 9 percent at a daily average temperature of
\(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm
rainfall a corresponding increase of the infection rate by approximately
2 percent.

@@

\section{Discussion}\label{discussion}

\subsection{Main findings}\label{main-findings}

@@

\subsection{Comparison with previous
research}\label{comparison-with-previous-research}

@@

In addition, in alignment with other recent published studies, our
results confirm evidence which associated a negative effect of
temperature on new COVID-19 cases
\citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}.
It is however controversial to other scientific literature describing no
effects
\citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020}
or even converse correlations
\citep{auler_evidence_2020, xie_association_2020}. The conflicting
results might be explained by different climates and characteristics of
the populations under study. While we are confident that our strict
causal analysis resulted in effect estimates as undistorted as possible,
there might be unconsidered bias in those other studies. Further
research needs to be done to elucidate the biological characteristics of
the novel virus SARS-CoV-2 regarding its ambient temperature survival
and transmission. In regards of humidity, our findings (higher humidity
leading to increased case numbers) are in agreement with previous
research \citep{qi_covid-19_2020, wu_effects_2020, juni_impact_2020}.
Finally, we found a positive causal effect of increment precipitation
and a raise in COVID-19 cases, which supports previous observations
\citep{sobral_association_2020}.

\subsection{Limitations and strengths}\label{limitations-and-strengths}

While use of a causal DAG is itself a strong tool to identify causal
effects (and not just statistical associations), it introduces two
limitations: causal assumptions within the graph (depicted by edges)
need to be well justified, and the statistical regression model that
calculates total causal effects needs to be appropriate for the task at
hand. We endorse our graph as a basis for discussion on residual
confounding.

We observed overdispersion and a substantial increase in model
performance with a negative binomial regression compared to Poisson
regression, which is in line with the results on COVID-19 daily case
counts of \citet{kraemer_effect_2020} and others
\citep{liu_impact_2020, bannister-tyrrell_preliminary_2020}. We did not
model case counts with a differential equation model like the classic
SIR-model \citep{kermack1991contributions} and its successors, since
these are more suited to prediction \citep[e.g.][]{rkimodellierung}
while our choice of a negative binomial regression framework allowed us
to estimate the effects of confounders more reliably. There are more
advanced statistical methods for count data, e.g.~zero-inflated models
and mixed models. We tested both approaches as extensions to the
negative binomial regression and experienced numerical problems and
increased computing time, along with an insubstantial increase in model
performance. Furthermore, our model assumed that all variables have
effects proportional to the size of their measurements. It is possible
that some variables show saturation effects or opposite effects for
lower, medium, or high values. This could be modelled with polynomial or
other transformations of the variables, which we did not employ due to
limited temporal and spatial data availability.

As we were using administrative data for our analysis, the results are
susceptible to the Modifiable Area Unit Problem (MAUP)
\citep{Openshaw1984}. The MAUP postulates that different regional
aggregations of the units of observation may lead to different results
and conclusions. Due to limited available data for the different
variables, there is currently no way to overcome this.

@@

\subsection{Discussion of causal
effects}\label{discussion-of-causal-effects}

In our analysis, the adverse effect of mobility in retail/recreation and
the favorable effect of mobility in grocery/pharmacy make a strong case
for the successful implementation of policies like contact restrictions
which limit the number of individual interactions that lead to
infections. This is due to retail/recreational areas encompassing mostly
places of social gatherings like restaurants and bars, malls, sports and
music venues, among others, while if people are doing more of their
essential shopping at supermarkets, they will most likely stay at home
with less contact to other people.

@@

\subsection{Conclusions}\label{conclusions}

To the best of our knowledge, this is the most comprehensive analysis of
causes for COVID-19 infections which integrates different data sources
(all publicly available). Causal reasoning with a DAG allows us to
estimate the causal effects reliably.

>>>>

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{causal-model}{%
\section{Causal Model}\label{causal-model}}

@@

A number of studies report a strong association of \textbf{mobility} restrictions on the number of new COVID-19 cases: Restrictive measures (e.g.~``stay-at-home'' orders, travel bans, or school closures) are shown to possibly reduce the COVID-19 incidence \citep{chang_modeling_2020, Chinazzi395, fowler_effect_2020, kraemer_effect_2020, lasry_timing_2020, linka_outbreak_2020, mazzoli_effects_2020, xiong_data-driven_2020}. However, some studies point out the combination of various non-pharmaceutical interventions (NPIs) is decisive to prevent new infections \citep{juni_impact_2020, lai_effect_2020}.

Google Trends \citep{google_trends} data can be used as a tool to get insights into public interest (\textbf{awareness}) in the coronavirus disease. Several recent studies imply a connection of relative search volumes (RSV) indices and reported new COVID-19 cases \citep{ayyoubzadeh_predicting_2020, effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, lin_google_2020, mavragani_tracking_2020, walker_use_2020, yuan_trends_2020, zhou_effects_2020}. Some search terms e.g.~``COVID-19'' or ``coronavirus'' predated newly infected cases/total number of cases by roughly 7 to 14 days for different countries \citep{effenberger_association_2020, higgins_correlations_2020, li_retrospective_2020, yuan_trends_2020}. Additionally, we acknowledged that individual risk-aware behavior might be a reaction to the current COVID-19 burden (measured as reported cases at the day of exposure).

@@

When investigating causal determinants of SARS-CoV-2 infections, a number of confounders have to be considered. Well-known risk factors for SARS-CoV-2 as well as for other infections are demographic factors such as age, gender, socio-economic status (SES), population density, and foreign citizenship/ethnicity \citep{de2020risk, Dragano2020.06.17.20133918, jhucovid19db2020}. In Germany along with other countries (i.e.~Brazil, USA, or the UK), populist parties or politicians and their electorate tend to be more sceptical about effects of containment measures than the other part of the electorate \citep{dohle_wingen_schreiber_2020, engle_staying_2020}. Therefore we considered both ``right-wing populist party votes'' and ``voter turnout'' as possible confounders. Public health interventions were also taken into account (contact restrictions, school closures etc.), as their implementation showed strong correlations with controlling the spread of SARS-CoV-2 \citep{cowling2020impact, juni_impact_2020, lai_effect_2020}. To avoid bias due to reporting delay of case numbers we had to include weekday and German holidays. We include some unobserved variables in our DAG (e.g.~``Herd immunity''), too. Please note that ``Exposure to SARS-CoV-2'' is itself an unobserved variable: German case numbers are reported with delay after date of exposure and symptom onset. \emph{Exposure to the virus} should not be confused with the formal \emph{exposure variables} of the DAG (mobility, awareness, temperature).

\hypertarget{data}{%
\section{Data}\label{data}}

@@

\hypertarget{variables}{%
\subsection{Variables}\label{variables}}

@@

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{causal-analysis-with-dag-and-adjustment-sets}{%
\subsection{Causal analysis with DAG and adjustment sets}\label{causal-analysis-with-dag-and-adjustment-sets}}

@@

\hypertarget{regression-with-negative-binomial-model}{%
\subsection{Regression with negative binomial model}\label{regression-with-negative-binomial-model}}

We can estimate the causal effect from exposure to outcome by regression \citep{pearl2009causality}. Since the outcome ``Reported new cases of COVID-19'' is a count variable, one should not employ a linear regression model with Gaussian errors, but instead we assumed a log-linear relationship between the expected value of the outcome \(Y\) (new cases) and regressors \(x\), as well as a Poisson or negative binomial distribution for \(Y\):
\begin{equation}
\log(\mathbb{E}[Y|x]) = \alpha + \sum_{i\in S}\beta_i\cdot x_{i}, \label{eq:regmodel}
\end{equation}
where \(\alpha\) is the regression intercept, \(S\) is the set of adjustment variables for the exposure \(i^{\ast}\) including the exposure variable itself, \(\beta_i\) are the regression coefficients corresponding to the variables \(x_i\). As such \(\beta_{i^{\ast}}\) is the total causal effect from exposure variable \(x_{i^{\ast}}\) on the outcome Y.

@@

We needed to account for the fact that our outcome is not counted per time unit (one day) only, but depends on the number of active COVID-19 cases: Holding all other variables fixed, the number of new cases \(Y\) is a constant proportion of the number of active cases \(A\). This was modeled by including an offset \(\log(A+1)\) in the regression model \eqref{eq:regmodel}:
\begin{align}
\log(\mathbb{E}[Y|x]) &= \alpha + \log(A+1) + \sum_{i\in S}\beta_i\cdot x_{i} \nonumber \\
\Leftrightarrow \log(\frac{\mathbb{E}[Y|x]}{A+1}) &= \alpha + \sum\beta_i\cdot x_{i} \label{eq:logratemodel} \\
\Leftrightarrow \frac{\mathbb{E}[Y|x]}{A+1} &= \exp(\alpha)\cdot \prod \exp(\beta_i)^{x_i}. \label{eq:ratechange}
\end{align}
Here we added a pseudocount ``\(+1\)'' to ensure a finite logarithm and avoid division by \(0\).

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{main-results}{%
\subsection{Main results}\label{main-results}}

@@

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.8 percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.5 percent. Mobility on workplaces showed a small effect of 0.3 increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 23 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 7 percent.

@@

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.8 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent. We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 9 percent at a daily average temperature of \(0^{\circ}C\). For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 2 percent.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{main-findings}{%
\subsection{Main findings}\label{main-findings}}

@@

\hypertarget{comparison-with-previous-research}{%
\subsection{Comparison with previous research}\label{comparison-with-previous-research}}

@@

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases \citep{bannister-tyrrell_preliminary_2020, demongeot_temperature_2020, liu_impact_2020, qi_covid-19_2020, shi_impact_2020, sobral_association_2020, tosepu_correlation_2020, Wang2020temperature, wu_effects_2020}. It is however controversial to other scientific literature describing no effects \citep{briz-redon_spatio-temporal_2020, iqbal_nexus_2020, jahangiri_sensitivity_2020, juni_impact_2020, yao_no_2020} or even converse correlations \citep{auler_evidence_2020, xie_association_2020}. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations \citep{sobral_association_2020}.

\hypertarget{limitations-and-strengths}{%
\subsection{Limitations and strengths}\label{limitations-and-strengths}}

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf.~\citep{gencoglu2020causal}). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g.~between the policies like contact restrictions and mandatory face masks). Another way to identify potential problems in the proposed DAG is to perform a sensitivity analysis of its structure by inspecting its maximal ancestral graph (MAG) or its Markov equivalence class represented by a complete partially DAG (CPDAG) and the existence of valid adjustment sets for these generalized graphs \citep{perkovic2017complete}. For the MAG derived from our DAG, only the effects for exposures mobility and searches for corona can be estimated with valid adjustment sets, while for the Markov equivalence class all exposures but COVID-19 burden lead to valid adjustments sets. A further analysis of these implications is out of the scope of this paper.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of \citet{kraemer_effect_2020} and others \citep{liu_impact_2020, bannister-tyrrell_preliminary_2020, Islamm2743}. We did not model case counts with a differential equation model like the classic SIR-model \citep{kermack1991contributions} and its successors, since these are more suited to prediction \citep[e.g.~][]{rkimodellierung} while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g.~zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers (similar to \citep{Islamm2743}).

We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level. As such, conclusions about effects cannot be transferred on individuals without the possibility for an ecological fallacy. Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) \citep{Openshaw1984}. The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions. Due to limited available data for the different variables, there is currently no way to overcome these problems that are inherent to all analyses on aggregated data level.

@@

\hypertarget{discussion-of-causal-effects}{%
\subsection{Discussion of causal effects}\label{discussion-of-causal-effects}}

In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers. This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.

@@

\hypertarget{conclusions}{%
\subsection{Conclusions}\label{conclusions}}

To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available). Causal reasoning with a DAG allows us to estimate the causal effects more reliably.
